Natural Stories model analyses

### 2021-08-19 Analysis/models.Rmd
Roger added bootstrapping to GAMs

@@

gam1 <- read_rds(here("Analysis/models/gam_predictions.rds")) %>%  ggplot( aes(x=surprisal, y=rt, ymin=CI_lower, ymax=CI_upper))+
  geom_line()+
  geom_ribbon(alpha=.3)+
  #facet_grid(s~model, scales="free_y")+
  facet_grid(s~model)+
  coord_cartesian(xlim=c(0,28))+
  labs(x="Surprisal (bits)", y="Reaction Time (ms)")+theme(axis.ticks.x=element_blank(), axis.title.x=element_blank(), axis.text.x=element_blank(), plot.margin=margin(t=0,r=0,b=0,l=0,unit="pt"))

>>>>

```{r bootstrap code}
require(OpenImageR)
gam_boot <- function(dat,formula,data_frames_for_prediction,N=100,method="REML",...) {
  models <- lapply(1:N,function(seed) {
    set.seed(seed)
    #dat[[weights_name]] <- tabulate(sample(1:nrow(dat),floor(nrow(dat)/5),replace=TRUE),nbins=nrow(dat))
    dat$mgcv_weights <- tabulate(sample(1:nrow(dat),nrow(dat),replace=TRUE),nbins=nrow(dat))
    #print(head(dat$mgcv_weights))
    #print(head(dat[[weights_name]]))
    #m <- gam(formula,data=dat,weights=dat[[weights_name]],...)
    m <- gam(formula,data=dat,weights=mgcv_weights,method=method,...)
    #print(m$deviance)
    return(m)
  })
  results <- lapply(names(data_frames_for_prediction),function(term) {
    predictions <-sapply(models,function(m)
      predict(m,newdata=data_frames_for_prediction[[term]],type="terms",terms=term))
    #print(dim(predictions))
    return(predictions)
  })
  #print(length(results))
  names(results) <- names(data_frames_for_prediction)
  return(results)
}

summarize_gam_boot_results <- function(gam_boot_results,data_frames_for_prediction,term_names,model_name,point_estimation_function=mean,quantiles=c(0.025,0.975)) {
  result <- lapply(1:length(data_frames_for_prediction),function(i) {
    #print(term_names[[i]])
    res <- tibble(x=data_frames_for_prediction[[i]][[term_names[[i]]]],
                  rt=apply(gam_boot_results[[i]],1,point_estimation_function),
                  CI_lower=apply(gam_boot_results[[i]],1,function(x) quantile(x,0.1)),
                  CI_upper=apply(gam_boot_results[[i]],1,function(x) quantile(x,0.9)),
                  model=model_name,
                  term=term_names[[i]])
    names(res)[names(res)=="x"] <- term_names[[i]]
    return(res)
  })
  return(result)
}

dat_for_surprisal_prediction <- expand_grid(surprisal=seq(0,25,by=0.01),freq=0,len=0,prev_surp=0,prev_freq=0,prev_len=0)
dat_for_freq_prediction <- expand_grid(surprisal=0,freq=seq(-9.5,7,by=0.01),len=0,prev_surp=0,prev_freq=0,prev_len=0)
dat_for_length_prediction <- expand_grid(surprisal=0,freq=0,len=seq(-3.4,10.6,by=0.01),prev_surp=0,prev_freq=0,prev_len=0)
dat_for_prev_surp_prediction <- expand_grid(surprisal=0,freq=0,len=0,prev_surp=seq(0,25,by=0.01),prev_freq=0,prev_len=0)
dat_for_prev_freq_prediction <- expand_grid(surprisal=0,freq=0,len=0,prev_surp=0,prev_freq=seq(-9.5,7,by=0.01),prev_len=0)
dat_for_prev_len_prediction <- expand_grid(surprisal=0,freq=0,len=0,prev_surp=0,prev_freq=0,prev_len=seq(-3.4,10.6,by=0.01))
#predict(m,newdata=dat_for_surprisal_prediction,type="terms",terms='s(surprisal)')
#predict(m,newdata=dat_for_freq_prediction,type="terms",terms='ti(freq)')
term_names <- c("surprisal","freq","len","prev_surp","prev_freq","prev_len")
dats_for_prediction <- list(
  's(surprisal)'=dat_for_surprisal_prediction,
  'ti(freq)'=dat_for_freq_prediction,
  'ti(len)'=dat_for_length_prediction,
  's(prev_surp)'=dat_for_prev_surp_prediction,
  'ti(prev_freq)'=dat_for_prev_freq_prediction,
  'ti(prev_len)'=dat_for_prev_len_prediction)
N_replicates <- 101
res_ngram <- gam_boot(ngram_mean_data,formula_no_interact,dats_for_prediction,N=N_replicates)
bootstrapped_gam_predictions_ngram <- summarize_gam_boot_results(res_ngram,dats_for_prediction,term_names,"5-gram")
res_grnn <- gam_boot(grnn_mean_data,formula_no_interact,dats_for_prediction,N=N_replicates)
bootstrapped_gam_predictions_grnn <- summarize_gam_boot_results(res_grnn,dats_for_prediction,term_names,"GRNN")
res_txl <- gam_boot(txl_mean_data,formula_no_interact,dats_for_prediction,N=N_replicates)
bootstrapped_gam_predictions_txl <- summarize_gam_boot_results(res_txl,dats_for_prediction,term_names,"Transformer-XL")
res_gpt <- gam_boot(gpt_mean_data,formula_no_interact,dats_for_prediction,N=N_replicates)
bootstrapped_gam_predictions_gpt <- summarize_gam_boot_results(res_gpt,dats_for_prediction,term_names,"GPT-2")
all_bootstrapped_predictions <- list(ngram=bootstrapped_gam_predictions_ngram,
                                     grnn=bootstrapped_gam_predictions_grnn,
                                     txl=bootstrapped_gam_predictions_txl,
                                     gpt=bootstrapped_gam_predictions_gpt)
all_surprisal_predictions <- rbind(bootstrapped_gam_predictions_ngram[[1]],
                                   rename(bootstrapped_gam_predictions_ngram[[4]],surprisal=prev_surp),
                                   bootstrapped_gam_predictions_grnn[[1]],
                                   rename(bootstrapped_gam_predictions_grnn[[4]],surprisal=prev_surp),
                                   bootstrapped_gam_predictions_txl[[1]],
                                   rename(bootstrapped_gam_predictions_txl[[4]],surprisal=prev_surp),
                                   bootstrapped_gam_predictions_gpt[[1]],
                                   rename(bootstrapped_gam_predictions_gpt[[4]],surprisal=prev_surp)
                                   ) %>%
  mutate(s=ifelse(term=="surprisal","Current","Previous")) %>%
  select(-term)

write_rds(all_surprisal_predictions,"bootstrapped_GAM_surprisal_predictions.rds")
```

```{r example bootstrap results}
# as an example, plot effect for surprisal
summaries <- bootstrapped_gam_predictions_grnn
ggplot(summaries[[1]],aes(x=surprisal,y=rt)) + geom_line() +
  #geom_rug(aes(x=grnn_mean_data$surprisal),sides='b') +
  geom_line(aes(x=summaries[[1]]$surprisal,y=summaries[[1]]$CI_lower),linetype="dashed") +
  geom_line(aes(x=summaries[[1]]$surprisal,y=summaries[[1]]$CI_upper),linetype="dashed") +
  ylim(c(-200,600)) +
  xlab("surprisal") +
  ylab("RT contribution") +
  theme(aspect.ratio=1)
ggsave("/tmp/bootstrap.pdf")
```

```{r example non-bootstrap results}
m_no_bootstrap <- gam(formula_no_interact,data=grnn_mean_data)
plot(m_no_bootstrap,select=1,ylim=c(-200,600))
pdf("/tmp/no_bootstrap.pdf")
plot(m_no_bootstrap,select=1,ylim=c(-200,600))
dev.off()
```

@@

#gam1 <- read_rds(here("Analysis/models/gam_predictions.rds")) %>%  
gam1 <- read_rds(here("Analysis/bootstrapped_GAM_surprisal_predictions.rds")) %>%  
  ggplot( aes(x=surprisal, y=rt, ymin=CI_lower, ymax=CI_upper))+
  geom_line()+
  geom_ribbon(alpha=.3)+
  #facet_grid(s~model, scales="free_y")+
  facet_grid(s~model)+
  coord_cartesian(xlim=c(0,28))+
  labs(x="Surprisal (bits)", y="Reaction Time (ms)")+theme(axis.ticks.x=element_blank(), axis.title.x=element_blank(), axis.text.x=element_blank(), plot.margin=margin(t=0,r=0,b=0,l=0,unit="pt"))
